Handling Recursion in Generic Programming Using Closed Type Families

### 2018-07-29 lncs-paper.tex
listings

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{url}
\usepackage{amssymb}

@@

\newcommand{\s}{$\hphantom{~}$}
\newcommand{\ind}{\s\s\s\s}
\newcommand{\hs}{\hspace{0.06cm}}
\newcommand{\ths}{\hspace{0.01cm}}
\newcommand{\tths}{\hspace{0.003cm}}
\newcommand{\nths}{\hspace{-0.01cm}}
\newcommand{\nhs}{\hspace{-0.06cm}}
\newcommand{\vs}{\vspace{0.2cm}\\}

\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}\s}
\newcommand{\comp}{{\tt>\nhs>\nhs>}}
\newcommand{\kleisli}{{\tt>\nhs=\nhs>}}
\newcommand{\comps}{\comp\s}
\newcommand{\kleislis}{\kleisli\s}
\newcommand{\ann}{:\nhs:\s}
\newcommand{\ds}{.\nhs.\nhs.}

@@

Another known solution uses overlapping instances. This, usually unwelcome, Haskell extension complicates reasoning about the semantics of code. In particular, the program behavior becomes unstable, for it can be affected by any module defining more specific instances. Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way. Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
% -- this should go to Realted Work (if stays in the paper at all)

@@

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types. The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
\texttt{
\vs
\indent\textbf{type family} Code (a \ann *) \ann [[*]]
\vs
}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

Consider, for instance, a datatype of binary trees:
\texttt{
\vs
\indent\textbf{data} Tree a = Leaf a | Node (Tree a) (Tree a)
\vs
}
This datatype is isomorphic to the following code:
\texttt{
\vs
\indent\textbf{type instance} Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\vspace{0.2cm}
}

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt \textbf{data} NP (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
\tt\s\s Nil\s\s \ann NP f '[]\\
\tt\s\s (:*) \ann f x \ra NP f xs \ra NP f (x ': xs)
\vs
\tt \textbf{data} NS (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
\tt\s\s Z \ann f x \ra NS f (x ': xs)\\
\tt\s\s S \ann NS f xs \ra NS f (x ': xs)
\end{tabular}
\caption{Datatypes for $n$-ary sums and products.}
\label{fig:ns-np}
\end{figure}

@@

As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types. The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the index list. Therefore, \texttt{NP} is a modest abstraction over a heterogeneous list.

The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic. The index list is allowed to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.
Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analogue of \texttt{const}:
\texttt{
\vs
\indent\textbf{newtype} I (a \ann *)\ind\ind\s\s\nhs = I \{unI \ann a\}\\
\indent\textbf{newtype} K (a \ann *) (b \ann k) = K \{unK \ann a\}
\vs
}
If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list, while \texttt{K a} turns it into a homogeneous one, isomorphic to \texttt{[a]}. Here is an example value of type \texttt{NP I}:
\texttt{
\vs
\indent I 5 :* I True :* I 'x' :* Nil \ann NP I '[Int, Bool, Char]
\vspace{0.2cm}
}

We turn to the sum definition now. The constructor \texttt{S} of \texttt{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \texttt{Z} stores the payload of type \texttt{f x}. For example, the following chooses the third element of a sum:
\texttt{
\vs
\indent S (S (Z (I 5))) \ann NS I '[Char, Bool, Int, Bool]
\vs
}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.

With the \texttt{NS} and \texttt{NP} machinery at hand, SOP defines the \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} witnessing the isomorphism between a datatype and its generic representation:
\texttt{
\vs
\indent\textbf{type} Rep a = SOP I (Code a)
\vs
\indent\textbf{class} All SListI (Code a) \Ra Generic (a \ann *) \textbf{where}\\
\indent\s\s \textbf{type} Code a \ann [[*]]\\
\indent\s\s from \ann a \ra Rep a\\
\indent\s\s to\s\s\s \ann Rep a \ra a
\vs
}
The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code. The functions, \texttt{from} and \texttt{to}, perform a shallow conversion of the datatype topmost layer---they do not recursively translate the constructor arguments.

We leave out discussion of the \texttt{SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work. Although, we do use \texttt{All} constraint combinator (as in \texttt{All SListI}) in the following. Therefore, it is worth noting that \texttt{All} applies a particular constraint (e.g. \texttt{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. We reimplement this function using the SOP view:
\texttt{
\vs
\indent subterms \ann Generic a \Ra a \ra [a]\\
\indent subterms t = subtermsNS \$ unSOP \$ from t
\vs
\indent subtermsNS \ann NS (NP I) xss \ra [a]\\
\indent subtermsNS (S ns) = subtermsNS ns\\
\indent subtermsNS (Z np) = subtermsNP np
\vs
\indent subtermsNP \ann \fa a xs. NP I xs \ra [a]\\
\indent subtermsNP p (I y :* ys)\\
\indent\s\s | typeEq @a y \hs\hs\ths= witnessEq y : subtermsNP ys\\
\indent\s\s | otherwise\s\s\s\hs\hs\ths = subtermsNP ys\\
\indent subtermsNP \_ Nil = []
\vs
}
The function \texttt{subterms} translates the term to its representation, unwrapping the sum of products from \texttt{SOP}, and passes that to the auxiliary function \texttt{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \texttt{subtermsNP}.

The algorithm of \texttt{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \texttt{t}, otherwise skipping the element. We use GHC's \textsf{TypeAppications} extension to pass that type.

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \texttt{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\texttt{
\vs
\indent\textbf{class} Subterms a (xs \ann [*]) \textbf{where}\\
\indent\s\s subtermsNP \ann NP I xs \ra [a]
\vs
\indent\textbf{instance} Subterms a xs \Ra Subterms a (x ': xs) \textbf{where}\\
\indent\s\s subtermsNP (\_\s\s\s :* xs) \hs\ths= subtermsNP xs\\
\indent\textbf{instance} \{-\# OVERLAPS \#-\}\\
\indent\ind\ind\s Subterms a xs \Ra Subterms a (a ': xs) \textbf{where}\\
\indent\s\s subtermsNP (I x :* xs) = x : subtermsNP xs\\
\indent\textbf{instance} Subterms a '[] \textbf{where}\\
\indent\s\s subtermsNP \_ = []
\vs
}

To make the whole solution work, we need to propagate the constraints all the way through \texttt{subtermsNS} and \texttt{subterms} signatures:
\texttt{
\vs
\indent subterms\s\s\s \ann\hs (Generic a, All (Subterms a) (Code a))\\
\indent\ind\ind\s\s\s \Ra a \ra [a]\\
\indent subtermsNS \ann\hs All (Subterms a) xss\\
\indent\ind\ind\s\s\s \Ra NS (NP I) xss \ra [a]
\vspace{0.2cm}
}

@@

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. The main idea of the extension is that the equations for a \emph{closed type family} are disallowed outside its declaration. Under the extension, we can give the following definition of type-level equality:
\texttt{
\vs
\indent\textbf{type family} Equal a x \ann Bool \textbf{where}\\
\indent\s\s Equal a a = 'True\\
\indent\s\s Equal a x = 'False
\vs
}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.

\subsection{Solution to \texttt{subtermsNP} revised}
\label{subsec:solve-subtermnp}

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
\texttt{
\vs
\indent\textbf{class} Proof (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s witnessEq \ann b \ra Maybe a
\vs
\indent\textbf{instance} Proof 'False a b \textbf{where}\\
\indent\s\s witnessEq \_ \hs\nths= Nothing\\
\indent\textbf{instance} Proof 'True\s\s a a \textbf{where}\\
\indent\s\s witnessEq\s\s\s = Just
\vs
}

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq}, which abbreviates the heavy-weighted interface of \texttt{Proof}:
\texttt{
\vs
\indent\textbf{class}\ind Proof (Equal a b) a b \Ra ProofEq a b\\
\indent\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
\vs
}
Resulting implementation of \texttt{subtermsNP} resembles our first definition given in the previous section:
\texttt{
\vs
\indent subtermsNP \ann \fa a xs. All (ProofEq a) xs \Ra NP I xs \ra [a]\\
\indent subtermsNP (I (y \ann x) :* ys)\\
\indent\s\s = case witnessEq @(Equal a x) y of\\
\indent\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
\indent\ind\s\s  Nothing \ra subtermsNP  ys\\
\indent subtermsNP \_ Nil = []
\vs
}

@@

To complete the solution of the problem, the \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms}' task. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

@@

The function \texttt{show} is a common example of useful functions that traverse a datatype's recursive structure. It is known that this function can be defined in a generic way for an arbitrary datatype. De Vries and L\"{o}h define generic function \texttt{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view. We follow their implementation of \texttt{gshow} for the most part, but improve it in respect of handling recursion. The original \texttt{gshow} yields to the standard \texttt{show} generated through \texttt{\textbf{deriving} Show}, because it does not consult with recursion points to place parentheses. We eliminate this drawback.

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function. Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \texttt{gshow}. Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
\texttt{
\vs
\indent\textbf{class} CaseShow (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s caseShow' \ann b \ra String
\vs
\indent\textbf{instance} Show\s\s b \Ra CaseShow 'False a b \textbf{where}\\
\indent\s\s caseShow'\s\s\s = show\\
\indent\textbf{instance} GShow a \Ra CaseShow 'True\s\s a a \textbf{where}\\
\indent\s\s caseShow' t = "(" ++ gshow t ++ ")"
\vs
}
We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise a synonym for the matching function:
\texttt{
\vs
\indent caseShow \ann \fa a b. CaseRecShow a b \Ra b \ra String\\
\indent caseShow t = caseShow' @(Equal a b) @a t
\vspace{0.2cm}
}

The resulting function \texttt{gshow} is a subject of a number of constraints abbreviated by a \texttt{GShow} synonym:
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String
\vs
}
The function \texttt{gshow} employs metadata provided by \textsf{generics-sop}'s class \texttt{Has\-Datatype\-Info} to show the names of a datatype constructor and its record fields. The \textsf{generics-sop} library is able to derive this metadata automatically.
The function is also constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types.

We define \texttt{gshow} mutually recursive with \texttt{caseShow}. The full implemenentation of the function \texttt{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.

The function \texttt{gshow} can now be used to generically show data---for example, a value of type \texttt{Tree~Bool}; note that \texttt{Tree~a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \texttt{Generic} and \texttt{HasDatatypeInfo}.
\texttt{
\vs
\indent *Main> \textbf{let} tree = Node (Leaf True) (Leaf False)\\
\indent *Main> gshow tree
\vs
\indent "Node (Leaf True) (Leaf False)"
\vs
}
Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
\texttt{
\vs
\indent\textbf{instance} Show T \textbf{where}\\
\indent\s\s show = gshow
\vs
}
This is a consequence of \texttt{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \texttt{Show} constraint for all knots in the datatype structure.

@@

The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}. Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g. \texttt{Tree~a} from Section~\ref{sec:sop-view}). We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent. It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.

Assume we have a polymorphically recursive datatype \texttt{PolyRec~a} defined in terms of a type family \texttt{Poly}:
\texttt{
\vs
\indent data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
\vs
\indent type family Poly a where\\
\indent \ind  Poly Bool = Char\\
\indent \ind  Poly Char = Bool\\
\indent \ind  Poly a\ind = a
\vspace{0.2cm}
}

For managing polymorphic recursion in this datatype, we can write an analogue of the \texttt{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types. Since any datatype with type parameters \texttt{f a b c \ds} has kind \texttt{* \ra (* \ra (* \ra \ds))}, a \texttt{PolyEq} type family can be defined thus:
\texttt{
\vs
\indent type family PolyEq (a \ann k) (x \ann k) \ann Bool where\\
\indent \ind  PolyEq (f a) (g b) = PolyEq f g\\
\indent \ind  PolyEq a\ind\s a\ind\s = 'True\\
\indent \ind  PolyEq a\ind\s x\ind\s = 'False
\vspace{0.2cm}
}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{url}
\usepackage{amssymb}
\usepackage{listings}
\usepackage{fancybox}

\lstset{
basicstyle=\ttfamily,
xleftmargin=\parindent,
aboveskip=0.2cm,
belowskip=0.2cm,
basewidth=0.5em,
numberbychapter=false,
escapeinside={(/}{/)},
literate={::}{{\ann}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\ds}}2
{>>>}{{\comp}}2 {>=>}{{\kleisli}}2
}

@@

\newcommand{\nhs}{\hspace{-0.06cm}}

\newcommand{\Ra}{$\Rightarrow$}
\newcommand{\ra}{$\rightarrow$}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}}
\newcommand{\comp}{>\nhs>\nhs>}
\newcommand{\kleisli}{>\nhs=\nhs>}
\newcommand{\ann}{:\texttt{\nhs}:}
\newcommand{\ds}{.\nhs.\nhs.}

\newcommand{\K}[1]{\lstinline{#1}}

\newcommand{\lset}{\lstset{basicstyle=\normalsize\ttfamily}}

\makeatletter
\lst@AddToHook{OnEmptyLine}{\vspace{-0.2255cm}}
\makeatother

\makeatletter
\newenvironment{CenteredBox}{
\begin{Sbox}}{
\end{Sbox}\centerline{\parbox{\wd\@Sbox}{\TheSbox}}}
\makeatother

@@

Another known solution uses overlapping instances. This, usually unwelcome, Haskell extension complicates reasoning about the semantics of code. In particular, the program behavior becomes unstable, for it can be affected by any module defining more specific instances. Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
The overlap problem also strikes in the security setting when code is compiled as \K{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way. Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
% -- this should go to Realted Work (if stays in the paper at all)

@@

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types. The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
\begin{lstlisting}
type family Code (a :: *) :: [[*]]
\end{lstlisting}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

Consider, for instance, a datatype of binary trees:
\begin{lstlisting}
data Tree a = Leaf a | Node (Tree a) (Tree a)
\end{lstlisting}
This datatype is isomorphic to the following code:
\begin{lstlisting}
type family Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\end{lstlisting}

\begin{figure}[t]
\begin{CenteredBox}
{\lset
\begin{lstlisting}
data NP (f :: k -> *) (xs :: [k]) where
  Nil  :: NP f '[]
  (:*) :: f x -> NP f xs -> NP f (x ': xs)

data NS (f :: k -> *) (xs :: [k]) where
  Z :: f x -> NS f (x ': xs)
  S :: NS f xs -> NS f (x ': xs)
\end{lstlisting}}
\end{CenteredBox}
\caption{Datatypes for $n$-ary sums and products.}
\label{fig:ns-np}
\end{figure}

@@

As shown in Figure~\ref{fig:ns-np}, the datatypes \K{NS} for an $n$-ary sum and \K{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types. The encoding also holds an auxiliary type constructor \K{f} (typically, a functor) which is meant to be applied to every element of the index list. Therefore, \K{NP} is a modest abstraction over a heterogeneous list.

The definitions of \K{NS} and \K{NP} are kind polymorphic. The index list is allowed to contain types of arbitrary kind \K{k}, since \K{k} turns to \K{*} by applying the type constructor \K{f}.
Basic instantiations of type parameter \K{f} found in SOP are identity functor \K{I}, that is, a type-level equivalent for \K{id} function, and a constant functor \K{K}, an analogue of \K{const}:
\begin{lstlisting}
newtype I (a :: *)          = I {unI :: a}
newtype K (a :: *) (b :: k) = K {unK :: a}
\end{lstlisting}
If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[a]}. Here is an example value of type \K{NP I}:
\begin{lstlisting}
I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
\end{lstlisting}

We turn to the sum definition now. The constructor \K{S} of \K{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \K{Z} stores the payload of type \K{f x}. For example, the following chooses the third element of a sum:
\begin{lstlisting}
S (S (Z (I 5))) :: NS I '[Char, Bool, Int, Bool]
\end{lstlisting}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.

With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to} witnessing the isomorphism between a datatype and its generic representation:
\begin{lstlisting}
type Rep a = SOP I (Code a)

class All SListI (Code a) => Generic (a :: *) where
  type Code a :: [[*]]
  from :: a -> Rep a
  to   :: Rep a -> a
\end{lstlisting}
The sum of products type, \K{SOP f}, is a newtype-wrapper for \K{NS (NP f)}, and the structural representation \K{Rep} of a datatype \K{a} is a type synonym for a \K{SOP I} of \K{a}'s code. The functions, \K{from} and \K{to}, perform a shallow conversion of the datatype topmost layer---they do not recursively translate the constructor arguments.

We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work. Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following. Therefore, it is worth noting that \K{All} applies a particular constraint (e.g. \K{SListI} above) to each member of a list of types. The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.

@@

We illustrate the problem through a short example. The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \K{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. We reimplement this function using the SOP view:
\begin{lstlisting}
subterms :: Generic a => a -> [a]
subterms t = subtermsNS $ unSOP $ from t

subtermsNS :: NS (NP I) xss -> [a]
subtermsNS (S ns) = subtermsNS ns
subtermsNS (Z np) = subtermsNP np

subtermsNP :: ∀a xs. NP I xs -> [a]
subtermsNP p (I y :* ys)
  | typeEq @a y = witnessEq y : subtermsNP ys
  | otherwise   = subtermsNP ys
subtermsNP _ Nil = []
\end{lstlisting}
The function \K{subterms} translates the term to its representation, unwrapping the sum of products from \K{SOP}, and passes that to the auxiliary function \K{subtermsNS}. The latter merely traverses the sum and, once reaches the product,  passes it further to \K{subtermsNP}.

The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \K{t}, otherwise skipping the element. We use GHC's \textsf{TypeAppications} extension to pass that type.

Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type. There is no clear path to this at the moment. Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
\begin{lstlisting}
class Subterms a (xs :: [*]) where
  subtermsNP :: NP I xs -> [a]

instance Subterms a xs => Subterms a (x ': xs) where
  subtermsNP (_   :* xs) = subtermsNP xs
instance {-# OVERLAPS #-}
         Subterms a xs => Subterms a (a ': xs) where
  subtermsNP (I x :* xs) = x : subtermsNP xs
instance Subterms a '[] where
  subtermsNP _ = []
\end{lstlisting}

To make the whole solution work, we need to propagate the constraints all the way through \K{subtermsNS} and \K{subterms} signatures:
\begin{lstlisting}
subterms   :: (Generic a, All (Subterms a) (Code a))
           => a -> [a]
subtermsNS :: All (Subterms a) xss
           => NS (NP I) xss -> [a]
\end{lstlisting}

@@

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. The main idea of the extension is that the equations for a \emph{closed type family} are disallowed outside its declaration. Under the extension, we can give the following definition of type-level equality:
\begin{lstlisting}
type family Equal a x :: Bool where
  Equal a a = 'True
  Equal a x = 'False
\end{lstlisting}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.

\subsection{Solution to \K{subtermsNP} revised}
\label{subsec:solve-subtermnp}

We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
\begin{lstlisting}
class Proof (eq :: Bool) (a :: *) (b :: *) where
  witnessEq :: b -> Maybe a

instance Proof 'False a b where
  witnessEq _ = Nothing
instance Proof 'True  a a where
  witnessEq   = Just
\end{lstlisting}

For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term. This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{Proof}:
\begin{lstlisting}
class    Proof (Equal a b) a b => ProofEq a b
instance Proof (Equal a b) a b => ProofEq a b
\end{lstlisting}
Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
\begin{lstlisting}
subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
subtermsNP (I (y :: x) :* ys)
  = case witnessEq @(Equal a x) y of
      Just t  -> t : subtermsNP ys
      Nothing -> subtermsNP  ys
subtermsNP _ Nil = []
\end{lstlisting}

@@

To complete the solution of the problem, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.

In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \K{subterms}' task. We give another example showing how to adapt our idiom to different scenarios in the following subsection.

@@

The function \K{show} is a common example of useful functions that traverse a datatype's recursive structure. It is known that this function can be defined in a generic way for an arbitrary datatype. De Vries and L\"{o}h define generic function \K{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view. We follow their implementation of \K{gshow} for the most part, but improve it in respect of handling recursion. The original \K{gshow} yields to the standard \K{show} generated through \K{deriving Show}, because it does not consult with recursion points to place parentheses. We eliminate this drawback.

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \K{Show}, and invoke its \K{show} function. Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \K{gshow}. Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
\begin{lstlisting}
class CaseShow (eq :: Bool) (a :: *) (b :: *) where
  caseShow' :: b -> String

instance Show  b => CaseShow 'False a b where
  caseShow'   = show
instance GShow a => CaseShow 'True  a a where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}
We provide a synonym for the \K{CaseShow (Equal a b) a b} instance, which we call \K{CaseRecShow}, as before with \K{ProofEq}; likewise a synonym for the matching function:
\begin{lstlisting}
caseShow :: ∀a b. CaseRecShow a b => b -> String
caseShow t = caseShow' @(Equal a b) @a t
\end{lstlisting}

The resulting function \K{gshow} is a subject of a number of constraints abbreviated by a \K{GShow} synonym:
\begin{lstlisting}
type GShow a = (Generic a, HasDatatypeInfo a,
                All2 (CaseRecShow a) (Code a))

gshow :: ∀a. GShow a => a -> String
\end{lstlisting}
The function \K{gshow} employs meta-information provided by \textsf{generics-sop}'s class \K{HasDatatypeInfo} to show the names of a datatype constructor and its record fields. The \textsf{generics-sop} library is able to derive this metadata automatically.
The function is also constrained by \K{CaseRecShow} with the \K{All2} combinator that is an analogue of \K{All} for a list of lists of types.

We define \K{gshow} mutually recursive with \K{caseShow}. The full implemenentation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.

The function \K{gshow} can now be used to generically show data---for example, a value of type \K{Tree Bool}; note that \K{Tree a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \K{Generic} and \K{HasDatatypeInfo}.
\begin{lstlisting}
*Main> let tree = Node (Leaf True) (Leaf False)
*Main> gshow tree

"Node (Leaf True) (Leaf False)"
\end{lstlisting}
Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}:
\begin{lstlisting}
instance Show T where
  show = gshow
\end{lstlisting}
This is a consequence of \K{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \K{Show} constraint for all knots in the datatype structure.

@@

The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}. Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g. \K{Tree a} from Section~\ref{sec:sop-view}). We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent. It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.

Assume we have a polymorphically recursive datatype \K{PolyRec a} defined in terms of a type family \K{Poly}:
\begin{lstlisting}
data PolyRec a = Tail a | Rec a (PolyRec (Poly a))

type family Poly a where
  Poly Bool = Char
  Poly Char = Bool
  Poly a    = a
\end{lstlisting}

For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types. Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
\begin{lstlisting}
type family PolyEq (a :: k) (x :: k) :: Bool where
  PolyEq (f a) (g b) = PolyEq f g
  PolyEq a     a     = 'True
  PolyEq a     x     = 'False
\end{lstlisting}

The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}. The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
\begin{lstlisting}
instance Show  b => CaseShow 'False a b where
  caseShow'   = show
instance GShow b => CaseShow 'True  a b where
  caseShow' t = "(" ++ gshow t ++ ")"
\end{lstlisting}
